---
title: "Vowel plots with `ggplot2`"
author: "Elise Bell"
date: "5/17/2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# These are the R packages you need to run this code correctly.
require(ggplot2)
require(ggpubr)
require(plyr)
require(dplyr)
require(reshape2)
```

## Read in data from a CSV

Minimally, your data should have the following columns:

> `[1] "vowel" "F1"    "F2"    "duration"`


You should also include columns for any additional information you have about your data (for example, tone or phonemic length) or subjects (group, age, etc.)

```{r}
data <- read.csv("pseudo_vowel_data.csv", header=T, encoding = "UTF-8")
```

Check that all vowel symbols were read in correctly. Use regular expressions and gsub() to correct any mis-read values, if necessary.
```{r}
unique(data$vowel)
data$vowel <- gsub("schwa", "ə", data$vowel, fixed=TRUE) # fixed=TRUE means that characters should be interpreted literally, not using regex interpretations
```

Make sure that the vowel column acts as a factor, not as character strings (do the same for any other categorical variables in your data set).
```{r}
data$vowel <- as.factor(data$vowel)
```

Look at the first six rows to make sure that everything looks okay:
```{r}
head(data)
```

Use `summary()` to get a summary of the entire dataframe, or any single column:
```{r}
summary(data$vowel)
```

## Creating Plots

Assuming we want to plot a F1 ~ F2 vowel space, we first need to tell `ggplot2` what data to look at
```{r}
ggplot(data=data)
```

We also need to tell it what variables it should pay attention to:
```{r}
ggplot(data=data, aes(x=F2, y=F1, label=vowel))
```

If you run that code, you get a plot, but no data appears, until you tell ggplot what to plot using a `geom_()` command
```{r}
ggplot(data=data, aes(x=F2, y=F1)) +
  geom_point()
```

Now we have data, but no information about what vowels it's from. You can specify that `color` should be used to indicate levels of the vowel factor
```{r}
ggplot(data=data, aes(x=F2, y=F1)) +
  geom_point(aes(color=vowel))
```

If your vowel data is non-normalized F1 and F2 values, you may notice that this plot does not match up with a traditional vowel plot (ex: /u/ should be in the top right). The way to fix this (if necessary) is to reverse the x and y axes. You can also re-position the axes so that the x-axis is on the top and the y-axis is on the right.

```{r}
ggplot(data=data, aes(x=F2, y=F1)) +
  geom_point(aes(color=vowel)) +
  scale_x_reverse() +
  scale_y_reverse()

ggplot(data=data, aes(x=F2, y=F1)) +
  geom_point(aes(color=vowel)) +
  scale_x_reverse(position = "top") +
  scale_y_reverse(position = "right")
```

We can also indicate vowel identity using `shape` instead of `color`.
```{r}
ggplot(data=data, aes(x=F2, y=F1)) +
  geom_point(aes(shape=vowel))
```

We can also plot the vowel symbols, using `geom_text()` instead of `geom_point()`.
```{r}
ggplot(data=data, aes(x=F2, y=F1)) +
  geom_text(aes(label=vowel))
```

You can combine `geom_text()` and `color`.
```{r}
ggplot(data=data, aes(x=F2, y=F1)) +
  geom_text(aes(label=vowel, color=vowel))
```

You may want to include both raw and summarized data. To summarize your data, you can use the `reshape2` package (for pipes `%>%`) and `dplyr`. This command produces a new dataframe (actually a tibble, don't worry about that) with one value for combination of vowel and v_length. It summarizes across speakers and across any other variables that aren't included in the `group_by()` command. Here, I only calculate means, but you can make many other summary calculations if desired.
```{r}
data.means <- data %>%
  group_by(vowel) %>%
  summarise(F1 = mean(F1),
            F2 = mean(F2),
            duration = mean(duration))
```

Once you have that summarized, you can add it to the plot containing the unsummarized data.
```{r}
ggplot(data=data, aes(x=F2, y=F1)) +
  geom_text(aes(label=vowel, color=vowel)) +
  geom_text(data=data.means, aes(label=vowel), size=12)
```

You can vary the transparency of layers of data by setting `alpha` to a value between 0 and 1. The smaller `alpha` is, the more transparent the result.
```{r}
ggplot(data=data, aes(x=F2, y=F1)) +
  geom_text(aes(label=vowel, color=vowel)) +
  geom_text(data=data.means, alpha=0.5, aes(label=vowel), size=12)
```

You can use the non-summarized data to create ellipses to represent the distribution of vowel groups.
```{r}
ggplot(data=data, aes(x=F2, y=F1)) +
  stat_ellipse(linetype=2, geom="polygon", alpha=0.25, aes(fill=vowel)) +
  geom_point(alpha=0.25, aes()) +
  geom_text(data=data.means, aes(label=vowel), size=12)
```

The plots above show two continuous variables (F1, F2) and each level of a categorical variable (vowel). If you only have one continuous variable (such as duration) and one categorical variable, you can create other types of plots.
```{r}
ggplot(data=data, aes(x=vowel, y=duration)) +
  geom_boxplot()
```

You can represent additional categorical variables (here, vowel length) in several different ways. You can show them all on the same plot using `color`.
```{r}
ggplot(data=data, aes(x=vowel, y=duration, color=v_length)) +
  geom_boxplot()
```

You can create a plot for each level of the variable using `facet_grid()`.
```{r}
ggplot(data=data, aes(x=vowel, y=duration, color=v_length)) +
  geom_boxplot() +
  facet_grid(.~v_length)
ggplot(data=data, aes(x=vowel, y=duration, color=v_length)) +
  geom_boxplot() +
  facet_grid(vowel~v_length)
```
  
##Other useful ggplot commands.

To change the theme of the plot, you can add a `theme_()` command.
```{r}
ggplot(data=data, aes(x=F2, y=F1)) +
  geom_point(aes(color=vowel)) +
  #theme_bw() 
  #theme_light()
  theme_void()
```

You can also modify the colors of the plot using `RColorBrewer`.
```{r}
require(RColorBrewer)
ggplot(data=data, aes(x=F2, y=F1,color=vowel)) +
  geom_point() +
  theme_bw() +
  scale_fill_brewer(palette="Dark2", aesthetics = c("color"))
```

To hide a legend that you don't want, try adding `show.legend=FALSE` to the `geom_` call.
```{r}
ggplot(data=data, aes(x=vowel, y=duration)) + geom_boxplot(aes(color=v_length))
ggplot(data=data, aes(x=vowel, y=duration)) + geom_boxplot(aes(color=v_length), show.legend=FALSE)
```

To change the font size of axis and legend text, use the `theme()` command.
```{r}
ggplot(data=data, aes(x=vowel, y=duration)) + 
  geom_boxplot(aes(color=v_length)) +
  theme(axis.text=element_text(size=24),
      axis.title=element_text(size=24),
      strip.text=element_text(size=24),
      legend.title=element_text(size=24),
      legend.text=element_text(size=24))
```

To change the default names of the x and y axes add `ylab()` and `xlab()`:
```{r}
ggplot(data=data, aes(x=vowel, y=duration)) + 
  geom_boxplot(aes(color=v_length)) +
  ylab("Duration (ms)") + 
  xlab("VOWELS")
```

To specify the limits of the axes, add limits to the call to `scale_x_continuous()`. (NB: `scale` commands have 3 parts, scale_, axis identifier (x, y), and type of axis (continuous, discrete, log, etc.))
```{r}
ggplot(data=data, aes(x=vowel, y=duration)) + 
  geom_boxplot(aes(color=v_length)) +
  scale_y_continuous(limits = c(0,500))
```

You can specify the title of the plot using `ggtitle()`.
```{r}
ggplot(data=data, aes(x=vowel, y=duration, color=v_length)) +
  geom_boxplot() +
  ggtitle("Vowel duration")
```

There are many many more things you can do with ggplot2 to display your data. Other tutorials you might find helpful are [here](http://joeystanley.com/blog/making-vowel-plots-in-r-part-1) and [here](http://www.mattwinn.com/tools/HB95_2.html). 
